Supermartingales and Termination Analysis of Statistical PCF

### 2020-05-27 probabilisticProgrammingMartingales.tex
Comment about Example (5) removed

\begin{proof}
\emph{Claim}: for all $n \in \omega$,
\(
\expect{Y_n} \leq \expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big).
\)
We prove the claim by induction on $n$.
The base case is trivial.
For the inductive case, we have
\iffalse
\everymath{\displaystyle}
\begin{calculation}
\int_A\nu(x,\dif y)\;\frac{\dif \nu'}{\dif\mu}(x,y)  
  \step{decomposition of $\nu = \kact{\mu }{ f}$} 
\int_A\mu(x,\dif y)\;f(x,y)\cdot \frac{\dif \nu'}{\dif\mu}(x,y)
  \step{$f=1$ outside $\infty[\nu]$} 
\int_A\mu(x,\dif y)\;\frac{\dif \nu'}{\dif\mu}(x,y)
  \step{RN Theorem for $\sigma$-finite measures}
\int_A\nu'(x,\dif y)
  \step{definition of integral}
\nu'(x,A).
\end{calculation}
\fi
{%\everymath{displaystyle}
\begin{calculation}
\expect{Y_{n+1}}
  \step[=]{definition of conditional expectation}
\expect{\expect{Y_{n+1} \mid \calF_n}}
  \step[\leq]{linearity of expectation}
\expect{Y_n} - \int \epsilon \circ Y_n \cdot 1_{\set{Y_n > 0}} \, \dif \mu
\; = \;
\expect{Y_n} - \int_{\set{Y_n > 0}} \epsilon \circ Y_n \, \dif \mu
  \step[\leq]{convexity of $\epsilon$, and Jensen's inequality}
\expect{Y_n} - \epsilon \big(\int_{\set{Y_n > 0}} Y_n \, \dif \mu\big)
  \step[\leq]{by assumption $Y_n \geq 0$}
\expect{Y_n} - \epsilon (\expect{Y_n})
  \step[\leq]{$\mathbb{P}[Y_n > 0] \leq 1$}
\expect{Y_n} - \epsilon (\expect{Y_n}) \cdot \mathbb{P}[Y_n > 0]
  \step[\leq]{$\epsilon$ is antitone, and $\epsilon_0 = \expect{Y_0} \geq \expect{Y_n}$}
\expect{Y_n} - \epsilon_0 \cdot \mathbb{P}[Y_n > 0]
  \step[\leq]{induction hypothesis}
%\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big) - \epsilon_0 \cdot \mathbb{P}[Y_n > 0]
%\step[\leq]{}
\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n} \mathbb{P}[Y_i > 0]\Big)
\end{calculation}
}

@@

For $M_1$, consider the function $g_1(n) := \frac{2^n - 1}{2^{n-1}}$.
Note that $(g_1 (n))_{n \geq 0}$ satisfies the recurrence relation
\[
z_0 = 0, \qquad
z_n > 1/2 \cdot z_{n-1} - 1/2 \cdot z_{n+1} \quad (n \geq 1)
\]
and set the antitone and convex function $\epsilon_1$ to be $x \mapsto g_1(x) - 1/2 \cdot g_1(x-1) - 1/2 \cdot g_1(x+1) = \frac{1}{2^{x+1}}$.

For $M_3$, we use the function $g_3(n) := \frac{3^n - 1}{3^{n-1}}$, and set $\epsilon_3$ to be $x \mapsto g_3(x) - 2/3 \cdot g_3(x-1) - 1/3 \cdot g_3(x+2) = 
\frac{28}{3^{x+2}}$.

@@

\lo{Example~\ref{ex:0 probability reachable} seems somewhat artificial: its validity depends on $I$ being the closed interval. 
There is no reason why we can't define $I$ to be the open interval $(0, 1)$, in which case, $Rch(M)$ would consist of all terms that are reachable from $M$ with non-zero probability.}
\end{enumerate}

>>>>

\begin{proof}
\emph{Claim}: for all $n \in \omega$,
\(
\expect{Y_n} \leq \expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big).
\)
We prove the claim by induction on $n$.
The base case is trivial.
For the inductive case, we have
\iffalse
\everymath{\displaystyle}
\begin{calculation}
\int_A\nu(x,\dif y)\;\frac{\dif \nu'}{\dif\mu}(x,y)  
  \step{decomposition of $\nu = \kact{\mu }{ f}$} 
\int_A\mu(x,\dif y)\;f(x,y)\cdot \frac{\dif \nu'}{\dif\mu}(x,y)
  \step{$f=1$ outside $\infty[\nu]$} 
\int_A\mu(x,\dif y)\;\frac{\dif \nu'}{\dif\mu}(x,y)
  \step{RN Theorem for $\sigma$-finite measures}
\int_A\nu'(x,\dif y)
  \step{definition of integral}
\nu'(x,A).
\end{calculation}
\fi
{%\everymath{displaystyle}
\begin{calculation}
\expect{Y_{n+1}}
  \step[=]{definition of conditional expectation}
\expect{\expect{Y_{n+1} \mid \calF_n}}
  \step[\leq]{linearity of expectation}
\expect{Y_n} - \int \epsilon \circ Y_n \cdot 1_{\set{Y_n > 0}} \, \dif \mu
\; = \;
\expect{Y_n} - \int_{\set{Y_n > 0}} \epsilon \circ Y_n \, \dif \mu
  \step[\leq]{convexity of $\epsilon$, and Jensen's inequality}
\expect{Y_n} - \epsilon \big(\int_{\set{Y_n > 0}} Y_n \, \dif \mu\big)
  \step[=]{by assumption $Y_n \geq 0$}
\expect{Y_n} - \epsilon (\expect{Y_n})
  \step[\leq]{$\mathbb{P}[Y_n > 0] \leq 1$}
\expect{Y_n} - \epsilon (\expect{Y_n}) \cdot \mathbb{P}[Y_n > 0]
  \step[\leq]{$\epsilon$ is antitone, and $\epsilon_0 = \expect{Y_0} \geq \expect{Y_n}$}
\expect{Y_n} - \epsilon_0 \cdot \mathbb{P}[Y_n > 0]
  \step[\leq]{induction hypothesis}
%\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[Y_i > 0]\Big) - \epsilon_0 \cdot \mathbb{P}[Y_n > 0]
%\step[\leq]{}
\expect{Y_0} - \epsilon_0 \cdot \Big(\sum_{i=0}^{n} \mathbb{P}[Y_i > 0]\Big)
\end{calculation}
}

@@

For $M_1$, consider the function $g_1(n) := \frac{2^n - 1}{2^{n-1}}$.
Note that $(g_1 (n))_{n \geq 0}$ satisfies the recurrence relation
\[
z_0 = 0, \qquad
z_n > 1/2 \cdot z_{n-1} - 1/2 \cdot z_{n+1} \quad (n \geq 1)
\]
and set the antitone and convex function $\epsilon_1$ to be 
\[
x \mapsto \big(g_1(x) - 1/2 \cdot g_1(x-1) - 1/2 \cdot g_1(x+1)\big) = \frac{1}{2^{x+1}}.
\]

For $M_3$, we use the function $g_3(n) := \frac{3^n - 1}{3^{n-1}}$, and set $\epsilon_3$ to be 
\[
x \mapsto \big(g_3(x) - 2/3 \cdot g_3(x-1) - 1/3 \cdot g_3(x+2) \big) = 
\frac{28}{3^{x+2}}.
\]

@@

%\lo{Example~\ref{ex:0 probability reachable} seems somewhat artificial: its validity depends on $I$ being the closed interval. There is no reason why we can't define $I$ to be the open interval $(0, 1)$, in which case, $Rch(M)$ would consist of all terms that are reachable from $M$ with non-zero probability.}
\end{enumerate}
